Structured Versus Decorated Cospans

### 2021-04-20 cospans_new.tex
a few inline comments

@@

We  prove Theorems \ref{thm:decorated_cospans_1} and \ref{thm:decorated_cospans_2} by applying this lemma to a particular monoidal bifibration.  To get our hands on this starting from our lax monoidal pseudofunctor $F \maps (A,+) \to (\Cat,\times)$, we need a monoidal version of the Grothendieck construction \cite{MV}.   We recall the Grothendieck construction in \ref{subsec:fibrations}; briefly,
this gives a bijection between pseudofunctors $F \maps \A \to \Cat$ and opfibrations
$U \maps \inta F \to \A$.   Moeller and the third author \cite{MV} showed that
similarly, given a (symmetric) monoidal category $\A$, there is a bijection between 
\begin{itemize}
\item \define{(symmetric) monoidal opindexed categories}: (symmetric) lax monoidal pseudofunctors \hfill \break $F\maps (\A, \otimes_A) \to (\Cat, \times)$
\end{itemize}
and 
\begin{itemize}
\item \define{(symmetric) monoidal opfibrations}: opfibrations $U \maps \X \to \A$ where $\X$ and $\A$ are (symmetric) monoidal, $U$ is (symmetric) strict monoidal, and $\otimes_\X$ preserves cocartesian liftings. 
\end{itemize}  
If the monoidal structure on $\A$ is in fact cocartesian, as it is in our case, there is a further correspondence between these structures and ordinary pseudofunctors $F\maps \A \to \MonCat$ (or to $\SMC$ in the symmetric case).

@@

\begin{proof}[Proof of \cref{thm:decorated_cospans_1}] 
To obtain the desired double category we apply \cref{lem:shulman} to a bifibration $\Phi \maps \A/U \to \A$ which we now define.   Let $U \maps \inta F \to \A$ be the monoidal opfibration associated to the lax monoidal pseudofunctor $F \maps \A \to \Cat$.   Thus:
\begin{itemize}
\item an object of $\inta F$ is a pair $(b,p)$ where $b \in \A$ and $p \in F(A)$;
\item a morphism in $\inta F$ from $(b,p)$ to $(b',p')$ is a pair $(g,\zeta)$
where $g \maps b \to b'$ is a morphism in $\A$ and $\zeta \maps F(f)(p) \to p'$ is
a morphism in $F(b')$.
\end{itemize}
The functor $U$ maps objects and morphisms to their first component.   We thus obtain an explicit description of the comma category $\A/U$:
\begin{itemize}
\item an object of $\A/U$ is a triple $(a,x,a \xrightarrow{i} U(x))$ with $a \in \A$ and $x=(b,p) \in \inta F$;
\item a morphism in $\A/U$ from $(a,x,a \xrightarrow{i} U(x))$ to $(a',x',a' \xrightarrow{i'} U(x'))$ is a pair $(f,\theta)$ where $f \maps a \to a'$ is a morphism in $\A$ and $\theta \maps x \to x'$ is a morphism in $\inta F$ such that the following square commutes:
\[
\begin{tikzpicture}[scale=1.5]
\node (A) at (0,0) {$a$};
\node (B) at (1.5,0) {$a'$};
\node (C) at (0,-1) {$U(x)$};
\node (D) at (1.5,-1) {$U(x')$.};
\path[->,font=\scriptsize,>=angle 90]
(A) edge node[above]{$f$} (B)
(A) edge node[left]{$i$} (C)
(B) edge node[right]{$i'$} (D)
(C) edge node[below]{$U(\theta)$} (D);
\end{tikzpicture}
\]
\end{itemize}
The functor $\Phi \maps \A/U \to \A$ is defined on objects by $\Phi(a,x,a \xrightarrow{i} U(x))=a$ and on morphisms by $\Phi(f,\theta)=f$. We have a monoidal structure on $\A/U$ where the tensor product of two objects $(a_1,x_1,a_1 \xrightarrow{i_1} U(x_1))$ and $(a_2,x_2,a_2 \xrightarrow{i_2} U(x_2))$ is given by
\[  (a_1,x_1,a_1 \xrightarrow{i_1} U(x_1)) \otimes (a_2,x_2,a_2 \xrightarrow{i_2} U(x_2)) = (a_1+a_2,x_1 \otimes x_2,a_1+a_2 \xrightarrow{i_1+i_2} U(x_1 \otimes x_2)) \]
where we have left the equality $U(x_1)+U(x_2) = U(x_1 \otimes x_2)$ implicit in the above composite, for reasons of space. The monoidal unit is given by $(0_\A,0_\X,0_\A \to U(0_\X))$. It is easy to see that the functor $\Phi$ is a strict monoidal using this tensor product.

@@

One can also check that the functor $\Phi$ is an opfibration, but this takes a bit more work.  Given a morphism $f \maps a \to a'$ in $\A$ and an object $(a,x,a \xrightarrow{i} U(x)) \in \A/U$ over $a$, we shall show that a cocartesian lifting of $f$ is given by 
\[ (f,\theta) \maps (a,x,a \xrightarrow{i} U(x)) \, \to \, (a',x', a' \xrightarrow{i'} U(x')). \]
Here $i'$ arises from the following pushout square in $\A$:
\[
\begin{tikzpicture}[scale=1.5]
\node (A) at (0,0) {$a$};
\node (B) at (1.5,0) {$a'$};
\node (C) at (0,-1) {$U(x)$};
\node (D) at (1.5,-1) {$U(x)+_a a'$};
\path[->,font=\scriptsize,>=angle 90]
(A) edge node[above]{$f$} (B)
(A) edge node[left]{$i$} (C)
(B) edge [dashed] node[right]{$i'$} (D)
(C) edge [dashed] node[below]{$g$} (D);
\end{tikzpicture}
\]
The object $x' \in \inta F$ is defined as follows.  Recall that $x \in \inta F$ is a
pair $(b,p)$ with $b\in \A$ and $p \in F(b)$, and note that $U(x) = b$.   Define 
$x'$ to be the pair $(b',p')$ where $b' = U(x)+_a a'$ is the pushout in the above
diagram and $p' = F(g)(p)$.  Then $U(x') = b'$ so we have 
$i' \maps a' \to U(x')$ as desired.  Finally, we define $\theta \maps x \to x'$ to be 
the morphism in $\inta F$ given by the pair $(g,\id)$ where $g \maps b \to b'$ is
given in the above diagram and $\id \maps F(f)(p) \to p'$ is the identity.

>>>>

{\chris I suggest to move the sentence about fibrations and bifibrations to the appendix.

A monoidal bifibration moreover has a global tensor product that preserves cartesian and cocartesian liftings.

I also suggest that the whole text introducing the Beck Chevalley condition for a bifibration is also moved to the appndix.}

@@

We  prove Theorems \ref{thm:decorated_cospans_1} and \ref{thm:decorated_cospans_2} by applying this lemma to a particular monoidal bifibration.  To get our hands on this starting from our lax monoidal pseudofunctor $F \maps (A,+) \to (\Cat,\times)$, we need a monoidal version of the Grothendieck construction \cite{MV}.   We recall the Grothendieck construction in \ref{subsec:fibrations}; briefly,
this gives a bijection between pseudofunctors $F \maps \A \to \Cat$ and opfibrations
$U \maps \inta F \to \A$.

{\chris It feels like there is some repetition in the material below and the sketch of the proof of 2.4. I believe the previous version of the paper had a briefer text here. No problem if you want to keep as is.}

Moeller and the third author \cite{MV} showed that
similarly, given a (symmetric) monoidal category $\A$, there is a bijection between 
\begin{itemize}
\item \define{(symmetric) monoidal opindexed categories}: (symmetric) lax monoidal pseudofunctors \hfill \break $F\maps (\A, \otimes_A) \to (\Cat, \times)$
\end{itemize}
and 
\begin{itemize}
\item \define{(symmetric) monoidal opfibrations}: opfibrations $U \maps \X \to \A$ where $\X$ and $\A$ are (symmetric) monoidal, $U$ is (symmetric) strict monoidal, and $\otimes_\X$ preserves cocartesian liftings. 
\end{itemize}  
If the monoidal structure on $\A$ is in fact cocartesian, as it is in our case, there is a further correspondence between these structures and ordinary pseudofunctors $F\maps \A \to \MonCat$ (or to $\SMC$ in the symmetric case).

@@

{\chris Some text saying that we have now set up the necessary framework to use Mike's result and obtain our double categories would be nice here}

\begin{proof}[Proof of \cref{thm:decorated_cospans_1}] 
To obtain the desired double category we apply \cref{lem:shulman} to a bifibration $\Phi \maps \A/U \to \A$ which we now define.   Let $U \maps \inta F \to \A$ be the monoidal opfibration associated to the lax monoidal pseudofunctor $F \maps \A \to \Cat$. {\chris Couldn't we just recall the appendix here?}  Thus:
\begin{itemize}
\item an object of $\inta F$ is a pair $(b,p)$ where $b \in \A$ and $p \in F(A)$;
\item a morphism in $\inta F$ from $(b,p)$ to $(b',p')$ is a pair $(g,\zeta)$
where $g \maps b \to b'$ is a morphism in $\A$ and $\zeta \maps F(f)(p) \to p'$ is
a morphism in $F(b')$.
\end{itemize}
The functor $U$ maps objects and morphisms to their first component.   We thus obtain an explicit description of the comma category $\A/U$ {\chris well the arrows are described like that in any slice category, there is nothing special for the Grothendieck category? Perhaps the slice description could be omitted or more brief?}:
\begin{itemize}
\item an object of $\A/U$ is a triple $(a,x,a \xrightarrow{i} U(x))$ with $a \in \A$ and $x=(b,p) \in \inta F$;
\item a morphism in $\A/U$ from $(a,x,a \xrightarrow{i} U(x))$ to $(a',x',a' \xrightarrow{i'} U(x'))$ is a pair $(f,\theta)$ where $f \maps a \to a'$ is a morphism in $\A$ and $\theta \maps x \to x'$ is a morphism in $\inta F$ such that the following square commutes:
\[
\begin{tikzpicture}[scale=1.5]
\node (A) at (0,0) {$a$};
\node (B) at (1.5,0) {$a'$};
\node (C) at (0,-1) {$U(x)$};
\node (D) at (1.5,-1) {$U(x')$.};
\path[->,font=\scriptsize,>=angle 90]
(A) edge node[above]{$f$} (B)
(A) edge node[left]{$i$} (C)
(B) edge node[right]{$i'$} (D)
(C) edge node[below]{$U(\theta)$} (D);
\end{tikzpicture}
\]
\end{itemize}
The functor $\Phi \maps \A/U \to \A$ is defined on objects by $\Phi(a,x,a \xrightarrow{i} U(x))=a$ and on morphisms by $\Phi(f,\theta)=f$. We have a monoidal structure on $\A/U$ where the tensor product of two objects $(a_1,x_1,a_1 \xrightarrow{i_1} U(x_1))$ and $(a_2,x_2,a_2 \xrightarrow{i_2} U(x_2))$ is given by
\[  (a_1,x_1,a_1 \xrightarrow{i_1} U(x_1)) \otimes (a_2,x_2,a_2 \xrightarrow{i_2} U(x_2)) = (a_1+a_2,x_1 \otimes x_2,a_1+a_2 \xrightarrow{i_1+i_2} U(x_1 \otimes x_2)) \]
where we have left the equality $U(x_1)+U(x_2) = U(x_1 \otimes x_2)$ implicit in the above composite, for reasons of space. The monoidal unit is given by $(0_\A,0_\X,0_\A \to U(0_\X))$. It is easy to see that the functor $\Phi$ is a strict monoidal using this tensor product.

@@

One can also check that the functor $\Phi$ is an opfibration, but this takes a bit more work. {\chris ...or ``is less straightforward'' to express more formally perhaps?} Given a morphism $f \maps a \to a'$ in $\A$ and an object $(a,x,a \xrightarrow{i} U(x)) \in \A/U$ over $a$, we shall show that a cocartesian lifting of $f$ is given by 
\[ (f,\theta) \maps (a,x,a \xrightarrow{i} U(x)) \, \to \, (a',x', a' \xrightarrow{i'} U(x')). \]
Here $i'$ arises from the following pushout square in $\A$:
\[
\begin{tikzpicture}[scale=1.5]
\node (A) at (0,0) {$a$};
\node (B) at (1.5,0) {$a'$};
\node (C) at (0,-1) {$U(x)$};
\node (D) at (1.5,-1) {$U(x)+_a a'$};
\path[->,font=\scriptsize,>=angle 90]
(A) edge node[above]{$f$} (B)
(A) edge node[left]{$i$} (C)
(B) edge [dashed] node[right]{$i'$} (D)
(C) edge [dashed] node[below]{$g$} (D);
\end{tikzpicture}
\]
The object $x' \in \inta F$ is defined as follows.  Recall that $x \in \inta F$ is a
pair $(b,p)$ with $b\in \A$ and $p \in F(b)$, and note that $U(x) = b$.   Define 
$x'$ to be the pair $(b',p')$ where $b' = U(x)+_a a'$ is the pushout in the above
diagram and $p' = F(g)(p)$.  Then $U(x') = b'$ so we have 
$i' \maps a' \to U(x')$ as desired.  Finally, we define $\theta \maps x \to x'$ to be 
the morphism in $\inta F$ given by the pair $(g,\id)$ where $g \maps b \to b'$ is
given in the above diagram and $\id \maps F(f)(p) \to p'$ is the identity.
